### 2017-11-14 submission/manuscript.Rmd
incorporate Will's method suggestions

***Diversity is Lower in Those with Carcinoma and Increases Relative Risk:***
Using power transformed and Z-score normalized alpha diversity metrics, both evenness and Shannon diversity in feces and not tissue were  lower in those with carcinoma [Figure 1]. Using linear mixed-effect models to control for study and variable region there was a significant decrease from control to adenoma to carcinoma for both evenness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))`) and Shannon diversity (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`). This effect was not observed in tissue when additionally controlling for whether the sample came from the same individual (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05). For fecal samples a decrease in Shannon diversity and evenness resulted in a significantly increased relative risk for carcinoma (P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))` and P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`, respectively) [Figure 2]. Although these values were significant, the effect size was relatively small for both metrics (Shannon RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` and evenness RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))`) [Figure 2]. Only a decrease in evenness had an increased relative risk for adenoma (P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2A & S1] but this effect size was even smaller than what was observed for carcinoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))`).

@@

***Obtaining Data Sets:***
Studies used for this meta-analysis were identified through the review articles written by Keku, et al. and Vogtmann, et al. [@keku_gastrointestinal_2015; @vogtmann_epidemiologic_2016]. Additional studies not mentioned in the reviews were obtained from the authors knowledge of the literature. All studies were included that used tissue or feces as their sample source for 16S rRNA gene sequencing analysis. Studies using either 454 or Illumina sequencing technology were included. Only data sets that had sequences available for analysis were included. Some studies did not have publically available sequences or did not have meta data in which the authors were able to share. After these filtering steps the following studies remained: Ahn [@ahn_human_2013], Baxter [@baxter_microbiota-based_2016], Brim [@brim_microbiome_2013], Burns [@burns_virulence_2015], Chen [@chen_human_2012], Dejea [@dejea_microbiota_2014], Flemer [@flemer_tumour-associated_2017], Geng [@geng_diversified_2013], Hale [@hale_shifts_2017], Kostic [@kostic_genomic_2012], Lu [@lu_mucosal_2016], Sanapareddy [@sanapareddy_increased_2012], Wang [@wang_structural_2012], Weir [@weir_stool_2013], and Zeller [@zeller_potential_2014]. The Zackular [@zackular_human_2014] study was not included becasue the 90 individuals analyzed within the study are contained within the larger Baxter study. The Kostic study was not used since after sequence processing all the case samples did not have more than 100 sequences remaining. This left a total of 14 studies in which complete analysis could be completed.

***Data Set Breakdown:***
In total there were 7 studies with only fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and Zeller), 5 studies with only tissue samples (Burns, Dejea, Geng, Lu, Sanapareddy), and 2 studies with both fecal and tissue samples (Chen and Flemer). The total number of individuals that were analyzed after sequence processing for feces was `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)` [Table 1]. The total number of matched and unmatched tissue samples that were analyzed after sequence processing was `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` [Table 2].

***Sequence Processing:***
For the majority of studies raw sequences were downloaded from the SRA (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata was obtained from the following website: http://www.ncbi.nlm.nih.gov/Traces/study/ by searching the respective accession number of the study. Of the studies that did not have sequences and meta data on the SRA one study had the data stored on DBGap [@ahn_human_2013] and for four studies the data was obtained directly from the authors [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013]. Each study was processed using the mothur (v1.39.3) software program [@schloss_introducing_2009]. Where possible quality filtering utilized the default methods used in mothur for either 454 or Illumina based sequencing. If it was not possible to use these defaults the author stated quality cut-offs were used instead. Chimeras were identifed and removed using the VSEARCH [@rognes_vsearch_2016] program and *de novo* OTU clustering at 97% similarity using the OptiClust algorithm [@westcott_opticlust_2017] was utilized.

***Statistical Analysis:***
All statistical analysis after sequence processing utilized the R software package (v`r paste(version$major, ".", version$minor, sep = "")`). For the alpha diversity analysis values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package. Testing for alpha diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package to correct for study and variable region effects in feces and study, variable region, and individual effects in tissue. Relative Risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages. Relative risk significance testing utilized the chi-squred test. Beta-diversity differences utilized a Bray-Curtis distance matrix and PERMANOVA executed with the vegan (v`r paste(packageVersion("vegan"))[1]`) package. Random Forest models were built using both the caret (v`r paste(packageVersion("caret"))[1]`) and randomForest (v`r paste(packageVersion("randomForest"))[1]`) packages. Random Forest testing of the obtained AUC versus a random model AUC utilized T-tests. Power analysis and estimations were made using the pwr (v`r paste(packageVersion("pwr"))[1]`) and statmod (v`r paste(packageVersion("statmod"))[1]`) packages. All figures were created using both ggplot2 (v`r paste(packageVersion("ggplot2"))[1]`) and gridExtra (v`r paste(packageVersion("gridExtra"))[1]`) packages.

***Study Analysis Overview:***
Alpha diversity was first assessed for differences between controls, adenoma, and carcinoma. We analyzed the data using linear mixed-effect models and relative risk. Beta-diversity was then assessed for each inidividual study. Next, four specific CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*) were assessed for differences in relative risk. We then built Random Forest models based on all genera or the select CRC-associated genera. The models were trained on one study then tested on the remaining studies for every study. The data was split between feces and tissue samples. Within the tissue groups the data was further divided between matched and unmatched tissue samples. Where applicable for each study predictions for adenoma and carcinoma were tested. This same approach was then applied at the OTU level with the exception that instead of testing on the other studies a 10-fold cross validation was utilized and 100 different models were created based on random 80/20 splitting of the data to generate a range of expected AUCs. For OTU based models the CRC associated genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*. The power of each study was assessed for an effect size ranging from 1% to 30%. An estimated sample n for these effect sizes was also generated based on 80% power.

***Reproducible Methods:***
The code and analysis can be found here https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017. Unless mentioned otherwise the accession number for the raw sequences for the studies used in this analysis can be found directly in the respective batch file, on the GitHub repository or in the original manuscript.

@@

**Figure 1: Alpha Diversity Differences between Control, Adenoma, and Carcinoma Across Sampling Site.** A) Alpha diversity metric differences by group in stool samples. B) Alpha diversity metric differences by group in unmatched tissue samples. C) Alpha diversity metric differences by group in matched tissue samples. The dashed line represents a Z-score of 0 or no difference from the median.

**Figure 2: Relative Risk for Adenoma or Carcinoma based on Alpha Diversity Metrics in Stool.** A) Alpha metric relative risk for adenoma. B) Alpha metric relative risk for carcinoma. Colors represent the different variable regions used within the respective study.

@@

**Figure S1: Relative Risk for Adenoma or Carcinoma based on Alpha Diversity Metrics in Tissue.** A) Alpha metric relative risk for adenoma. B) Alpha metric relative risk for carcinoma. Colors represent the different variable regions used within the respective study.

>>>>

***Diversity is Lower in Those with Carcinoma and Increases Relative Risk:***
Using power transformed and Z-score normalized $\alpha$-diversity metrics, both evenness and Shannon diversity in feces and not tissue were  lower in those with carcinoma [Figure 1]. Using linear mixed-effect models to control for study and variable region there was a significant decrease from control to adenoma to carcinoma for both evenness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))`) and Shannon diversity (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`). This effect was not observed in tissue when additionally controlling for whether the sample came from the same individual (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05). For fecal samples a decrease in Shannon diversity and evenness resulted in a significantly increased relative risk for carcinoma (P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))` and P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`, respectively) [Figure 2]. Although these values were significant, the effect size was relatively small for both metrics (Shannon RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` and evenness RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))`) [Figure 2]. Only a decrease in evenness had an increased relative risk for adenoma (P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2A & S1] but this effect size was even smaller than what was observed for carcinoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))`).

@@

***Obtaining Data Sets:***
Studies used for this meta-analysis were identified through the review articles written by Keku, *et al.* and Vogtmann, *et al.* [@keku_gastrointestinal_2015; @vogtmann_epidemiologic_2016]. Additional studies not mentioned in the reviews were obtained based on the authors' knowledge of the literature. Studies that used tissue or feces as their sample source for 16S rRNA gene sequencing analysis were included. Studies using either 454 or Illumina sequencing technology were included. Only data sets that had sequences available for analysis were included. Some studies did not have publically available sequences or did not have metadata in which the authors were able to share. After these filtering steps, the following studies remained: Ahn, *et al.* [@ahn_human_2013], Baxter, *et al.* [@baxter_microbiota-based_2016], Brim, *et al.* [@brim_microbiome_2013], Burns, *et al.* [@burns_virulence_2015], Chen, *et al.*  [@chen_human_2012], Dejea, *et al.* [@dejea_microbiota_2014], Flemer, *et al.* [@flemer_tumour-associated_2017], Geng, *et al.* [@geng_diversified_2013], Hale, *et al.* [@hale_shifts_2017], Kostic, *et al.* [@kostic_genomic_2012], Lu, *et al.* [@lu_mucosal_2016], Sanapareddy, *et al.* [@sanapareddy_increased_2012], Wang, *et al.* [@wang_structural_2012], Weir, *et al.* [@weir_stool_2013], and Zeller, *et al.*  [@zeller_potential_2014]. The Zackular [@zackular_human_2014] study was not included becasue the 90 individuals analyzed within the study are contained within the larger Baxter study [@baxter_microbiota-based_2016]. Additionally, after sequence processing all the case samples for the Kostic study only had 100 sequences remaining and was not used. This left a total of 14 studies for which analysis could be completed.

***Data Set Breakdown:***
In total, there were seven studies with only fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and Zeller), five studies with only tissue samples (Burns, Dejea, Geng, Lu, Sanapareddy), and two studies with both fecal and tissue samples (Chen and Flemer). The total number of individuals that were analyzed after sequence processing for feces was `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)` [Table 1]. The total number of matched and unmatched tissue samples that were analyzed after sequence processing was `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` [Table 2].

***Sequence Processing:***
For the majority of studies raw sequences were downloaded from the Sequence Read Archive (SRA) (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata was obtained from the by searching the respective accession number of the study following website: http://www.ncbi.nlm.nih.gov/Traces/study/. Of the studies that did not have sequences and metadata on the SRA, data was obtained from DBGap for one study [@ahn_human_2013] and for four studies was obtained directly from the authors [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013]. Each study was processed using the mothur (v1.39.3) software program [@schloss_introducing_2009]. Where possible, quality filtering utilized the default methods used in mothur for either 454 or Illumina based sequencing. If it was not possible to use these defaults, the stated quality cut-offs were used instead. Chimeras were identifed and removed using VSEARCH [@rognes_vsearch_2016] before *de novo* OTU clustering at 97% similarity using the OptiClust algorithm [@westcott_opticlust_2017] was utilized.

***Statistical Analysis:***
All statistical analysis after sequence processing utilized the R software package (v`r paste(version$major, ".", version$minor, sep = "")`). For the $\alpha$-diversity analysis, values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package. Testing for $\alpha$-diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package to correct for study and variable region effects in feces and study, variable region, and individual effects in tissue. Relative risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages. Relative risk significance testing utilized the chi-squred test. $\beta$-diversity differences utilized a Bray-Curtis distance matrix and PERMANOVA executed with the vegan (v`r paste(packageVersion("vegan"))[1]`) package. Random Forest models were built using both the caret (v`r paste(packageVersion("caret"))[1]`) and randomForest (v`r paste(packageVersion("randomForest"))[1]`) packages. Differences between the obtained AUC versus a random model AUC was assessed using T-tests. Power analysis and estimations were made using the pwr (v`r paste(packageVersion("pwr"))[1]`) and statmod (v`r paste(packageVersion("statmod"))[1]`) packages. All figures were created using both ggplot2 (v`r paste(packageVersion("ggplot2"))[1]`) and gridExtra (v`r paste(packageVersion("gridExtra"))[1]`) packages.

***Study Analysis Overview:***
$\alpha$-diversity was first assessed for differences between controls, adenoma, and carcinoma. We analyzed the data using linear mixed-effect models and relative risk. $\beta$-diversity was then assessed for each inidividual study. Next, four specific CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*) were assessed for differences in relative risk. We then built Random Forest models based on all genera or the select CRC-associated genera. The models were trained on one study then tested on the remaining studies for every study. The data was split between feces and tissue samples. Within the tissue groups the data was further divided between matched and unmatched tissue samples. Where applicable for each study, predictions for adenoma and carcinoma were tested. This same approach was then applied at the OTU level with the exception that instead of testing on the other studies, a 10-fold cross validation was utilized and 100 different models were created based on random 80/20 splitting of the data to generate a range of expected AUCs. For OTU based models the CRC associated genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*. The power of each study was assessed for an effect size ranging from 1% to 30%. An estimated sample n for these effect sizes was also generated based on 80% power.

***Reproducible Methods:***
The code and analysis can be found here https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017. Unless mentioned otherwise, the accession number for the raw sequences for the studies used in this analysis can be found directly in the respective batch file in the GitHub repository or in the original manuscript.

@@

**Figure 1: $\alpha$-Diversity Differences between Control, Adenoma, and Carcinoma Across Sampling Site.** A) $\alpha$-diversity metric differences by group in stool samples. B) $\alpha$-diversity metric differences by group in unmatched tissue samples. C) $\alpha$-diversity metric differences by group in matched tissue samples. The dashed line represents a Z-score of 0 or no difference from the median.

**Figure 2: Relative Risk for Adenoma or Carcinoma based on $\alpha$-Diversity Metrics in Stool.** A) $\alpha$-metric relative risk for adenoma. B) $\alpha$-metric relative risk for carcinoma. Colors represent the different variable regions used within the respective study.

@@

**Figure S1: Relative Risk for Adenoma or Carcinoma based on $\alpha$-Diversity Metrics in Tissue.** A) $\alpha$-metric relative risk for adenoma. B) $\alpha$-metric relative risk for carcinoma. Colors represent the different variable regions used within the respective study.

### 2017-11-14 submission/manuscript.tex
incorporate Will's method suggestions

@@

\textbf{\emph{Diversity is Lower in Those with Carcinoma and Increases
Relative Risk:}} Using power transformed and Z-score normalized alpha
diversity metrics, both evenness and Shannon diversity in feces and not
tissue were lower in those with carcinoma {[}Figure 1{]}. Using linear
mixed-effect models to control for study and variable region there was a
significant decrease from control to adenoma to carcinoma for both
evenness (P-value = 0.025) and Shannon diversity (P-value = 0.043). This
effect was not observed in tissue when additionally controlling for
whether the sample came from the same individual (P-value \textgreater{}
0.05). For fecal samples a decrease in Shannon diversity and evenness
resulted in a significantly increased relative risk for carcinoma
(P-value = 0.01 and P-value = 0.0011, respectively) {[}Figure 2{]}.
Although these values were significant, the effect size was relatively
small for both metrics (Shannon RR = 1.31 and evenness RR = 1.34)
{[}Figure 2{]}. Only a decrease in evenness had an increased relative
risk for adenoma (P-value = 0.032) {[}Figure 2A \& S1{]} but this effect
size was even smaller than what was observed for carcinoma (RR = 1.16).

@@

\textbf{\emph{Obtaining Data Sets:}} Studies used for this meta-analysis
were identified through the review articles written by Keku, et al. and
Vogtmann, et al. {[}26,27{]}. Additional studies not mentioned in the
reviews were obtained from the authors knowledge of the literature. All
studies were included that used tissue or feces as their sample source
for 16S rRNA gene sequencing analysis. Studies using either 454 or
Illumina sequencing technology were included. Only data sets that had
sequences available for analysis were included. Some studies did not
have publically available sequences or did not have meta data in which
the authors were able to share. After these filtering steps the
following studies remained: Ahn {[}21{]}, Baxter {[}24{]}, Brim
{[}17{]}, Burns {[}22{]}, Chen {[}14{]}, Dejea {[}19{]}, Flemer
{[}13{]}, Geng {[}25{]}, Hale {[}12{]}, Kostic {[}28{]}, Lu {[}16{]},
Sanapareddy {[}20{]}, Wang {[}15{]}, Weir {[}18{]}, and Zeller {[}23{]}.
The Zackular {[}29{]} study was not included becasue the 90 individuals
analyzed within the study are contained within the larger Baxter study.
The Kostic study was not used since after sequence processing all the
case samples did not have more than 100 sequences remaining. This left a
total of 14 studies in which complete analysis could be completed.

\textbf{\emph{Data Set Breakdown:}} In total there were 7 studies with
only fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and Zeller), 5
studies with only tissue samples (Burns, Dejea, Geng, Lu, Sanapareddy),
and 2 studies with both fecal and tissue samples (Chen and Flemer). The
total number of individuals that were analyzed after sequence processing
for feces was 1737 {[}Table 1{]}. The total number of matched and
unmatched tissue samples that were analyzed after sequence processing
was 492 {[}Table 2{]}.

\textbf{\emph{Sequence Processing:}} For the majority of studies raw
sequences were downloaded from the SRA
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained from the following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/} by searching the
respective accession number of the study. Of the studies that did not
have sequences and meta data on the SRA one study had the data stored on
DBGap {[}21{]} and for four studies the data was obtained directly from
the authors {[}12,13,18,20{]}. Each study was processed using the mothur
(v1.39.3) software program {[}30{]}. Where possible quality filtering
utilized the default methods used in mothur for either 454 or Illumina
based sequencing. If it was not possible to use these defaults the
author stated quality cut-offs were used instead. Chimeras were
identifed and removed using the VSEARCH {[}31{]} program and \emph{de
novo} OTU clustering at 97\% similarity using the OptiClust algorithm
{[}32{]} was utilized.

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R software package (v3.4.2). For the
alpha diversity analysis values were power transformed using the
rcompanion (v1.10.1) package and then Z-score normalized using the car
(v2.1.5) package. Testing for alpha diversity differences utilized
linear mixed-effect models created using the lme4 (v1.1.14) package to
correct for study and variable region effects in feces and study,
variable region, and individual effects in tissue. Relative Risk was
analyzed using both the epiR (v0.9.87) and metafor (v2.0.0) packages.
Relative risk significance testing utilized the chi-squred test.
Beta-diversity differences utilized a Bray-Curtis distance matrix and
PERMANOVA executed with the vegan (v2.4.4) package. Random Forest models
were built using both the caret (v6.0.77) and randomForest (v4.6.12)
packages. Random Forest testing of the obtained AUC versus a random
model AUC utilized T-tests. Power analysis and estimations were made
using the pwr (v1.2.1) and statmod (v1.4.30) packages. All figures were
created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages.

\textbf{\emph{Study Analysis Overview:}} Alpha diversity was first
assessed for differences between controls, adenoma, and carcinoma. We
analyzed the data using linear mixed-effect models and relative risk.
Beta-diversity was then assessed for each inidividual study. Next, four
specific CRC-associated genera (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, and \emph{Porphyromonas}) were assessed for
differences in relative risk. We then built Random Forest models based
on all genera or the select CRC-associated genera. The models were
trained on one study then tested on the remaining studies for every
study. The data was split between feces and tissue samples. Within the
tissue groups the data was further divided between matched and unmatched
tissue samples. Where applicable for each study predictions for adenoma
and carcinoma were tested. This same approach was then applied at the
OTU level with the exception that instead of testing on the other
studies a 10-fold cross validation was utilized and 100 different models
were created based on random 80/20 splitting of the data to generate a
range of expected AUCs. For OTU based models the CRC associated genera
included all OTUs that had a taxonomic classification to
\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus}, or
\emph{Porphyromonas}. The power of each study was assessed for an effect
size ranging from 1\% to 30\%. An estimated sample n for these effect
sizes was also generated based on 80\% power.

\textbf{\emph{Reproducible Methods:}} The code and analysis can be found
here
\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017}.
Unless mentioned otherwise the accession number for the raw sequences
for the studies used in this analysis can be found directly in the
respective batch file, on the GitHub repository or in the original
manuscript.

@@

\textbf{Figure 1: Alpha Diversity Differences between Control, Adenoma,
and Carcinoma Across Sampling Site.} A) Alpha diversity metric
differences by group in stool samples. B) Alpha diversity metric
differences by group in unmatched tissue samples. C) Alpha diversity
metric differences by group in matched tissue samples. The dashed line
represents a Z-score of 0 or no difference from the median.

\textbf{Figure 2: Relative Risk for Adenoma or Carcinoma based on Alpha
Diversity Metrics in Stool.} A) Alpha metric relative risk for adenoma.
B) Alpha metric relative risk for carcinoma. Colors represent the
different variable regions used within the respective study.

@@

\textbf{Figure S1: Relative Risk for Adenoma or Carcinoma based on Alpha
Diversity Metrics in Tissue.} A) Alpha metric relative risk for adenoma.
B) Alpha metric relative risk for carcinoma. Colors represent the
different variable regions used within the respective study.

>>>>

\usepackage{amsmath}

\usepackage{mathtools}

@@

\textbf{\emph{Diversity is Lower in Those with Carcinoma and Increases
Relative Risk:}} Using power transformed and Z-score normalized
\(\alpha\)-diversity metrics, both evenness and Shannon diversity in
feces and not tissue were lower in those with carcinoma {[}Figure 1{]}.
Using linear mixed-effect models to control for study and variable
region there was a significant decrease from control to adenoma to
carcinoma for both evenness (P-value = 0.025) and Shannon diversity
(P-value = 0.043). This effect was not observed in tissue when
additionally controlling for whether the sample came from the same
individual (P-value \textgreater{} 0.05). For fecal samples a decrease
in Shannon diversity and evenness resulted in a significantly increased
relative risk for carcinoma (P-value = 0.01 and P-value = 0.0011,
respectively) {[}Figure 2{]}. Although these values were significant,
the effect size was relatively small for both metrics (Shannon RR = 1.31
and evenness RR = 1.34) {[}Figure 2{]}. Only a decrease in evenness had
an increased relative risk for adenoma (P-value = 0.032) {[}Figure 2A \&
S1{]} but this effect size was even smaller than what was observed for
carcinoma (RR = 1.16).

@@

\textbf{\emph{Obtaining Data Sets:}} Studies used for this meta-analysis
were identified through the review articles written by Keku, \emph{et
al.} and Vogtmann, \emph{et al.} {[}26,27{]}. Additional studies not
mentioned in the reviews were obtained based on the authors' knowledge
of the literature. Studies that used tissue or feces as their sample
source for 16S rRNA gene sequencing analysis were included. Studies
using either 454 or Illumina sequencing technology were included. Only
data sets that had sequences available for analysis were included. Some
studies did not have publically available sequences or did not have
metadata in which the authors were able to share. After these filtering
steps, the following studies remained: Ahn, \emph{et al.} {[}21{]},
Baxter, \emph{et al.} {[}24{]}, Brim, \emph{et al.} {[}17{]}, Burns,
\emph{et al.} {[}22{]}, Chen, \emph{et al.} {[}14{]}, Dejea, \emph{et
al.} {[}19{]}, Flemer, \emph{et al.} {[}13{]}, Geng, \emph{et al.}
{[}25{]}, Hale, \emph{et al.} {[}12{]}, Kostic, \emph{et al.} {[}28{]},
Lu, \emph{et al.} {[}16{]}, Sanapareddy, \emph{et al.} {[}20{]}, Wang,
\emph{et al.} {[}15{]}, Weir, \emph{et al.} {[}18{]}, and Zeller,
\emph{et al.} {[}23{]}. The Zackular {[}29{]} study was not included
becasue the 90 individuals analyzed within the study are contained
within the larger Baxter study {[}24{]}. Additionally, after sequence
processing all the case samples for the Kostic study only had 100
sequences remaining and was not used. This left a total of 14 studies
for which analysis could be completed.

\textbf{\emph{Data Set Breakdown:}} In total, there were seven studies
with only fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and
Zeller), five studies with only tissue samples (Burns, Dejea, Geng, Lu,
Sanapareddy), and two studies with both fecal and tissue samples (Chen
and Flemer). The total number of individuals that were analyzed after
sequence processing for feces was 1737 {[}Table 1{]}. The total number
of matched and unmatched tissue samples that were analyzed after
sequence processing was 492 {[}Table 2{]}.

\textbf{\emph{Sequence Processing:}} For the majority of studies raw
sequences were downloaded from the Sequence Read Archive (SRA)
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained from the by searching the respective accession
number of the study following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
not have sequences and metadata on the SRA, data was obtained from DBGap
for one study {[}21{]} and for four studies was obtained directly from
the authors {[}12,13,18,20{]}. Each study was processed using the mothur
(v1.39.3) software program {[}30{]}. Where possible, quality filtering
utilized the default methods used in mothur for either 454 or Illumina
based sequencing. If it was not possible to use these defaults, the
stated quality cut-offs were used instead. Chimeras were identifed and
removed using VSEARCH {[}31{]} before \emph{de novo} OTU clustering at
97\% similarity using the OptiClust algorithm {[}32{]} was utilized.

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R software package (v3.4.2). For the
\(\alpha\)-diversity analysis, values were power transformed using the
rcompanion (v1.10.1) package and then Z-score normalized using the car
(v2.1.5) package. Testing for \(\alpha\)-diversity differences utilized
linear mixed-effect models created using the lme4 (v1.1.14) package to
correct for study and variable region effects in feces and study,
variable region, and individual effects in tissue. Relative risk was
analyzed using both the epiR (v0.9.87) and metafor (v2.0.0) packages.
Relative risk significance testing utilized the chi-squred test.
\(\beta\)-diversity differences utilized a Bray-Curtis distance matrix
and PERMANOVA executed with the vegan (v2.4.4) package. Random Forest
models were built using both the caret (v6.0.77) and randomForest
(v4.6.12) packages. Differences between the obtained AUC versus a random
model AUC was assessed using T-tests. Power analysis and estimations
were made using the pwr (v1.2.1) and statmod (v1.4.30) packages. All
figures were created using both ggplot2 (v2.2.1) and gridExtra (v2.3)
packages.

\textbf{\emph{Study Analysis Overview:}} \(\alpha\)-diversity was first
assessed for differences between controls, adenoma, and carcinoma. We
analyzed the data using linear mixed-effect models and relative risk.
\(\beta\)-diversity was then assessed for each inidividual study. Next,
four specific CRC-associated genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas})
were assessed for differences in relative risk. We then built Random
Forest models based on all genera or the select CRC-associated genera.
The models were trained on one study then tested on the remaining
studies for every study. The data was split between feces and tissue
samples. Within the tissue groups the data was further divided between
matched and unmatched tissue samples. Where applicable for each study,
predictions for adenoma and carcinoma were tested. This same approach
was then applied at the OTU level with the exception that instead of
testing on the other studies, a 10-fold cross validation was utilized
and 100 different models were created based on random 80/20 splitting of
the data to generate a range of expected AUCs. For OTU based models the
CRC associated genera included all OTUs that had a taxonomic
classification to \emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, or \emph{Porphyromonas}. The power of each
study was assessed for an effect size ranging from 1\% to 30\%. An
estimated sample n for these effect sizes was also generated based on
80\% power.

\textbf{\emph{Reproducible Methods:}} The code and analysis can be found
here
\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017}.
Unless mentioned otherwise, the accession number for the raw sequences
for the studies used in this analysis can be found directly in the
respective batch file in the GitHub repository or in the original
manuscript.

@@

\textbf{Figure 1: \(\alpha\)-Diversity Differences between Control,
Adenoma, and Carcinoma Across Sampling Site.} A) \(\alpha\)-diversity
metric differences by group in stool samples. B) \(\alpha\)-diversity
metric differences by group in unmatched tissue samples. C)
\(\alpha\)-diversity metric differences by group in matched tissue
samples. The dashed line represents a Z-score of 0 or no difference from
the median.

\textbf{Figure 2: Relative Risk for Adenoma or Carcinoma based on
\(\alpha\)-Diversity Metrics in Stool.} A) \(\alpha\)-metric relative
risk for adenoma. B) \(\alpha\)-metric relative risk for carcinoma.
Colors represent the different variable regions used within the
respective study.

@@

\textbf{Figure S1: Relative Risk for Adenoma or Carcinoma based on
\(\alpha\)-Diversity Metrics in Tissue.} A) \(\alpha\)-metric relative
risk for adenoma. B) \(\alpha\)-metric relative risk for carcinoma.
Colors represent the different variable regions used within the
respective study.
